Revisiting the hydrodynamic wave modulation from linear wave theory

### 2024-08-28 paper/paper.tex
Working on the text

\begin{equation}
\label{eq:lhs1960wavenumber}
\widetilde{k} = k (1 + \varepsilon_L)
\end{equation}
where $\widetilde{k}$ and $k$ are the modulated and unmodulated short-wave wavenumbers,
respectively, and $\varepsilon_L = a_L k_L$ is the steepness of the long wave with
wavenumber $k_L$ and amplitude $a_L$.
This result influenced the development of remote sensing of ocean
surface waves and currents \citep{keller1975microwave,hara1994hydrodynamic}.
\citet{phillips1981dispersion} and \citet{longuet1987propagation} revisited this
problem by nonlinear long waves and variations in the effective gravity of short
waves, yielding significantly stronger modulation than previously predicted.
\citet{henyey1988energy} derived an analytical solution for the modulation of
short waves by long waves using Hamiltonian mechanics and reported the modulation
of similar magnitude to that of \citet{longuet1987propagation}.
\citet{zhang1990evolution} considered weakly nonlinear short waves
using a non-linear Schr\" odinger equation and found stronger wavenumber
modulation but weaker amplitude modulation than those predicted by
\citet{longuet1987propagation}.
The variety of analytical and numerical frameworks (Eulerian, Hamiltonian,
Schr\" odinger) that relatively closely reproduce steady solutions for short-wave
modulation suggests that the solutions are robust.
Nevertheless, alternative analytical and numerical approaches to the modulation
problem are still of interest, especially if higher degrees of nonlinearity
can be considered using simpler approaches.

More recently, \citep{peureux2021unsteady} asked whether the steady solutions
of short-wave modulation are realistic by simulating the full wave conservation
equations
They found that the short waves grow unsteadily due to the propagation of
longer waves, suggesting that the steady-state solutions may only be valid for
a few long-wave periods before the short waves begin to break due to excessive
steepness.
However, the unsteadiness of their solutions occurs in a specific scenario of
a uniform short-wave field that suddenly appears in a uniform long-wave train,
and the numerical simulations stabilize if the short-wave field is initialized
from existing steady solutions.

@@

In this paper, we revisit this problem from a linear wave theory perspective and
derive new non-linear solutions for the modulation of short waves, evaluated
at the long-wave surface.
We also perform numerical simulations of the full wave crest and action
conservation laws and evaluate the effect of long-wave groups on the short waves.
The next section presents the governing equations, followed by the analytical
solutions in section \ref{section:analytical_solutions}.
In section \ref{section:numerical_solutions} we present the numerical solutions
and compare them to the analytical solutions.
In section \ref{section:scale_analysis} we perform a scale analysis of the
analytical solutions to quantify the relative importance of the terms in the
full equation set.
Finally, we summarize the results and discuss their implications in section
\ref{section:conclusions}.

@@

In this paper we consider inviscid, irrotational, and incompressible flow.
In these conditions the linear (sine-like), deep-water, surface gravity waves
obey the dispersion relation:

\begin{equation}
\label{eq:dispersion}
\omega = \sqrt{gk} + k U
\end{equation}
where $\omega$ is the angular frequency, $g$ is the gravitational acceleration,
$k$ is the wavenumber, and $U$ is the mean advective current in the direction
of the wave propagation.
A current in the direction of the waves thus increases its absolution frequency,
wavenumber remaining unchanged.
An important limitation here is that $U$ must be slow-varying on the scales of
the wave period \citep{bretherton1968wavetrains}.
In the context of short waves riding on long waves, the mean advective current
is the horizontal near-surface orbital velocity of the long wave.

The second important relation is the conservation of wave crests
\citep{phillips1981dispersion}:

@@

\begin{equation}
\label{eq:wave_action}
\dfrac{\partial N}{\partial t}
+ \dfrac{\partial}{\partial x} \left[\left(C_g + U\right)N\right]
= 0
\end{equation}
where $N$ is the action density, and $C_g = \partial \omega / \partial k$ is the
group speed.
In deep water, the group speed is half the phase speed, so $C_g = 1/2\sqrt{g/k}$.

As the short waves ride on the surface of the long wave,
they move in an accelerated reference frame and will thus experience effective
gravitational acceleration that changes in both space and time
\citep{longuet1986eulerian,longuet1987propagation}.
Inserting Eq. (\ref{eq:dispersion}) into (\ref{eq:wave_action}) we obtain:

\begin{equation}
\label{eq:wavenumber}
\dfrac{\partial k}{\partial t}
+ \left(C_g + U\right) \dfrac{\partial k}{\partial x}
+ k \dfrac{\partial U}{\partial x}
+ \dfrac{1}{2} \sqrt{\dfrac{k}{g}} \dfrac{\partial g}{\partial x}
= 0
\end{equation}
From left to right, the terms in this equation represent the change in wavenumber
due to the propagation and advection by ambient current $U$, the convergence of
the ambient current, and the spatial inhomogeneity of the gravitational acceleration.
In a non-accelerated reference frame (i.e. in absence of longer waves), the last
term is zero.
However, if gravitational acceleration varies in space, this term is necessary
to conserve the wave crests.

@@

\begin{equation}
\label{eq:wave_action2}
\dfrac{\partial N}{\partial t}
+ \left(C_g + U\right) \dfrac{\partial N}{\partial x}
+ N \dfrac{\partial U}{\partial x}
+ N \dfrac{\partial C_g}{\partial x}
= 0
\end{equation}
This equation is similar to Eq. \ref{eq:wavenumber}, except for the last term
which here represents the inhomogeneity of the group speed.
As \citet{peureux2021unsteady} explained, this term is responsible for unstable
growth of wave action in infinite long-wave trains.
Eqs. \ref{eq:wavenumber} and \ref{eq:wave_action2} will be the governing equations
that we solve numerically in section \ref{section:numerical_solutions}.

For the ambient forcing, we consider a train of monochromatic long wave
defined, to the first order in steepness, by the elevation $\eta_L = a_L cos{\psi}$,
where $a_L$ is the long-wave amplitude, $\psi = k_L x - \omega_L t$ is the
long-wave phase, and $\omega_L$ its angular frequency.

At this time we define the long-wave steepness $\varepsilon_L = a_L k_L$ which
we will use thoughout the rest of the paper as a scaling parameter.
The long wave induces horizontal and vertical orbital velocities whose values
at the surface $\eta_L$ are:

@@

Eqs. \ref{eq:dispersion}-\ref{eq:w_L} is the full system of equations that we
exploit in the remainder of this paper.
Note that here we derive the solutions from a different starting set of
equations than that of \citet{longuet1960changes}.
Rather than starting from the conservation of crests and wave action, as we do
here, they derived their solution from the linear superposition of the velocity
potential of two waves.
It appears the starting from the conservation of crests and wave action allows
for a simpler and more intuitive derivation.

>>>>

\begin{equation}
\label{eq:lhs1960wavenumber}
\widetilde{k} = k (1 + \varepsilon_L)
\end{equation}
where $\widetilde{k}$ and $k$ are the modulated and unmodulated short-wave wavenumbers,
respectively, and $\varepsilon_L = a_L k_L$ is the steepness of the long wave with
wavenumber $k_L$ and amplitude $a_L$.
This result influenced the development of remote sensing of ocean
surface waves and currents \citep{keller1975microwave,hara1994hydrodynamic}.
\citet{phillips1981dispersion} and \citet{longuet1987propagation} revisited this
problem by nonlinear long waves and variations in the effective gravity of short
waves, yielding significantly stronger modulation than previously predicted.
\citet{henyey1988energy} derived an analytical solution for the modulation of
short waves by long waves using Hamiltonian mechanics and reported the modulation
of similar magnitude to that of \citet{longuet1987propagation}.
\citet{zhang1990evolution} considered weakly nonlinear short waves
using a non-linear Schr\" odinger equation and found stronger wavenumber
modulation but weaker amplitude modulation than those predicted by
\citet{longuet1987propagation}.
The variety of analytical and numerical frameworks (Eulerian, Hamiltonian,
Schr\" odinger) that relatively closely reproduce steady solutions for short-wave
modulation suggests that the solutions are robust.
Alternative analytical and numerical approaches to the modulation problem remain
of interest, especially if higher degrees of nonlinearity can be considered
using simpler approaches.

More recently, \citep{peureux2021unsteady} asked whether the steady solutions
of short-wave modulation are realistic by simulating the full wave conservation
equations
They found that the short waves grow unsteadily due to the propagation of
longer waves, suggesting that the steady-state solutions may only be valid for
a few long-wave periods before the short waves begin to break due to excessive
steepness.
However, the unsteadiness of their solutions occurs in a specific scenario of
a uniform short-wave field that suddenly appears in a uniform long-wave train,
and the numerical simulations stabilize if the short-wave field is initialized
from existing steady solutions.
Nevertheless, their results do put into question whether the short-wave
modulation predicted by theory is generally steady over many long-wave periods
or not.

In this paper, we revisit this problem from linear wave theory and
derive alternative non-linear solutions for the modulation of short waves,
evaluated at the long-wave surface.
The analytical solutions yield modulations that are stronger than the original solution
by \citet{longuet1960changes}, but similar to the nonlinear numerical solutions
by \citet{longuet1987propagation} and \citet{zhang1990evolution} in long-wave
steepness up to $\approx 0.2$.
We also perform numerical simulations of the full wave crest and action
conservation laws and evaluate the effect of long-wave groups on the short waves.
The numerical simulations are used to determine the validity of the approximate
steady solution and to investigate the unsteadiness of the short-wave modulation
found by \citet{peureux2021unsteady}.
The main prior studies on hydrodynamic modulation, and their assumptions and
approaches to the solution, are summarized in Table
\ref{table:modulation_literature_summary}.

@@

This paper first presents the governing equations and then the analytical
solutions for the modulation of short waves by long waves and compares them to
the original solution by \citet{longuet1960changes}.
Then, we compare the analytical solutions to numerical solutions of the full
wave conservation equations in section \ref{section:numerical_solutions}.
In section \ref{section:scale_analysis} we perform a scale analysis of the
analytical solutions to quantify the relative importance of the terms in the
full equation set.
Finally, we summarize the results and discuss their implications in section
\ref{section:conclusions}.

@@

We consider inviscid, irrotational, and incompressible flow.
In these conditions the linear (sine-like), deep-water, surface gravity waves
obey the dispersion relation:

\begin{equation}
\label{eq:dispersion}
\omega = \sqrt{gk} + k U
\end{equation}
where $\omega$ is the angular frequency, $g$ is the gravitational acceleration,
$k$ is the wavenumber, and $U$ is the mean advective current in the direction
of the wave propagation.
A current in the direction of the waves increases its absolute frequency,
while the wavenumber remains unchanged.
An important limitation here is that $U$ must be slowly varying on the scales of
the wave period \citep{bretherton1968wavetrains}.
In the context of short waves riding on long waves, the mean advective current
is the horizontal near-surface orbital velocity of the long wave.
The evolution of the short-wave wavenumber is described by the conservation of
wave crests \citep{phillips1981dispersion}:

@@

\begin{equation}
\label{eq:wave_action}
\dfrac{\partial N}{\partial t}
+ \dfrac{\partial}{\partial x} \left[\left(C_g + U\right)N\right]
= 0
\end{equation}
where $N$ is the action density, and $C_g = \partial \omega / \partial k$ is the
group speed.
In deep water, the group speed is half the phase speed, so $C_g = 1/2\sqrt{g/k}$.
Although the absence of wave growth and dissipation is not generally realistic
in the open ocean, it is a helpful approximation that allows us to isolate the
hydrodynamic modulation effects solely due to the motion of the long waves.

Short waves that ride on the surface of the longer waves move in an accelerated
reference frame and thus experience effective gravitational acceleration that
changes in space and time \citep{phillips1981dispersion,longuet1986eulerian,longuet1987propagation}.
From Eqs. (\ref{eq:dispersion}) and (\ref{eq:wave_action}) we obtain:

\begin{equation}
\label{eq:wavenumber}
\dfrac{\partial k}{\partial t}
+ \left(C_g + U\right) \dfrac{\partial k}{\partial x}
+ k \dfrac{\partial U}{\partial x}
+ \dfrac{1}{2} \sqrt{\dfrac{k}{g}} \dfrac{\partial g}{\partial x}
= 0
\end{equation}
From left to right, the terms in this equation represent the change in wavenumber
due to the propagation and advection by ambient current $U$, the convergence of
the ambient current, and the spatial inhomogeneity of the gravitational acceleration.
In a non-accelerated reference frame (i.e. in absence of longer waves), the last
term vanishes.
Otherwise, this term is necessary to conserve the wave crests.

@@

\begin{equation}
\label{eq:wave_action2}
\dfrac{\partial N}{\partial t}
+ \left(C_g + U\right) \dfrac{\partial N}{\partial x}
+ N \dfrac{\partial U}{\partial x}
+ N \dfrac{\partial C_g}{\partial x}
= 0
\end{equation}
This equation is similar to Eq. \ref{eq:wavenumber}, except for the last term
which here represents the inhomogeneity of the group speed.
As \citet{peureux2021unsteady} explained, this term is responsible for unstable
growth of wave action in infinite long-wave trains.
Eqs. \ref{eq:wavenumber} and \ref{eq:wave_action2} are the governing equations
that we approximate and solve analytically in section \ref{section:analytical_solutions},
and solve numerically in their full form in section \ref{section:numerical_solutions}.

For the ambient forcing, we consider a train of monochromatic long wave
defined, to the first order in steepness, by the elevation $\eta_L = a_L cos{\psi}$,
where $a_L$ is the long-wave amplitude, $\psi = k_L x - \omega_L t$ is the
long-wave phase, and $\omega_L$ its angular frequency.
At this time we define the long-wave steepness $\varepsilon_L = a_L k_L$ which
we will use throughout the rest of the paper as a scaling parameter, as is common
in the literature.
The long wave induces horizontal and vertical orbital velocities whose values
at the surface $\eta_L$ are:

@@

Eqs. \ref{eq:dispersion}-\ref{eq:w_L} is the full system of equations that we
use in this paper.
Note that here we derive the solutions from a different starting set of
equations than that of \citet{longuet1960changes}.
Rather than starting from the conservation of crests and wave action, as we do
here, they derived their solution from the linear superposition of the velocity
potential of two waves.
It appears that starting from the conservation of crests and wave action allows
for a simpler and more intuitive derivation.
